---
title: "EDA.rmd"
author: "Carlo van Buiten"
date: "9/15/2021"
output: pdf_document
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(reshape2)
```

\newpage

# Research question
Is cell uniformity (shape and size) a good predictor for the malignancy of breast cancer?


In this dataset all attributes are grades ranging from 1 to 10, with 10 being the most severe. Each row is a unique case of breast-cancer. For specific percentages etc. refer to the user manual below:

Attributes definitions: https://www.rai-light.com/docs/BCD_User_Manual_v01.pdf


I hypothesize that it makes sense for a lower cell uniformity to coincide with a higher severity of (breast) cancer, since one the defining aspects of cancer is the erratic way the cells grow and multiply.




Let us start by taking a look at the data distribution.
```{r, warning=FALSE, message=FALSE}
data <- read.csv("breast-cancer-wisconsin.data", na.strings="?")
colnames(data) <- c("id", "Clump_Thickness", "Cell_Size_Uniformity", "Cell_Shape_Uniformity", "Marginal_Adhesion",
                    "Single_Epithelial_Cell_Size", "Bare_Nuclei", "Bland_Chromatin", "Normal_Nucleoli", "Mitoses",
                    "Class")

# head(data, 10)
summary(data)
```
We can see the data does not follow a normal distribution.


Let us take a look at the highest correlating features by only taking the data columns that correlate with malignancy for more than 80%.
```{r warning=FALSE, message=FALSE}
cor_data <- cor(data, use = "complete.obs")
cor_data <- as.data.frame(cor_data)
cor_cols <- names(cor_data)[which(cor_data$Class > 0.8)]
cor_cols 
```
There are 3 attributes that correlate with malignancy for more than 80%, 2 of which are pertaining to cell uniformity. This bodes well for the hypothesis that cell uniformity could potentially be a good predictor of malignant breast cancer.

Let us look at what the spreads of these attributes look like with regards to the 2 classes, where B is benign and M is malignant.

```{r, warning=FALSE, message=FALSE}
plot_data <- data
plot_data$Class[plot_data$Class == 2] <- "B"
plot_data$Class[plot_data$Class == 4] <- "M"

ggplot(data=plot_data, mapping = aes(x = Class, y = Cell_Size_Uniformity)) +
  geom_jitter(width = 0.05, height = 0.3, col="red", alpha = 0.3) +
  labs(y="Cell Size Uniformity", 
       caption="Figure 1: Cell size uniformity spread per class.") +
  theme_minimal() +
  theme(plot.caption.position = "plot",
        plot.caption = element_text(hjust = 0.5))
```

```{r, warning=FALSE, message=FALSE}
ggplot(data=plot_data, mapping = aes(x = Class, y = Cell_Shape_Uniformity)) +
  geom_jitter(width = 0.05, height = 0.3, col="green", alpha = 0.3) +
  labs(y="Cell Shape Uniformity", 
       caption="Figure 2: Cell shape uniformity spread per class.") +
  theme_minimal() +
  theme(plot.caption.position = "plot",
        plot.caption = element_text(hjust = 0.5))
```

```{r, warning=FALSE, message=FALSE}
ggplot(data=plot_data, mapping = aes(x = Class, y = Bare_Nuclei)) +
  geom_jitter(width = 0.05, height = 0.3, col="blue", alpha = 0.3) +
  labs(y="Nuclei With Cytoplasm", 
       caption="Figure 3: Bare nuclei spread per class.") +
  theme_minimal() +
  theme(plot.caption.position = "plot",
        plot.caption = element_text(hjust = 0.5))
```
It does seem to be the case that the more individual cells differ from each other (that is to say; being less uniform), the higher the chances are for the cancer to be malignant. The same goes for finding cytoplasm in the cell nucleus.
However, it looks like these 3 attributes correlate quite strongly not only with the class, but also with each other. It would make sense for cell shape and size uniformity to covariate, but let us actually compare each of these features with each other. 

```{r, warning=FALSE, message=FALSE}
ggplot(data=plot_data, mapping = aes(x = Cell_Size_Uniformity, y = Cell_Shape_Uniformity)) +
  geom_point(aes(col=Class)) +
  geom_smooth(method = "lm", se = T) +
  labs(y="Cell Shape Uniformity", x="Cell Size Uniformity", 
       caption="Figure 4: Plot of the correlation between cell shape and size discrepancies") +
  theme_minimal() +
  theme(plot.caption.position = "plot",
        plot.caption = element_text(hjust = 0.5))
```

Unsurprisingly, it does seem to be the case that the larger the differences in size between cells, the larger the differences in shapes as well. This makes sense as both these differences can be explained by erratic cell growth. Continuing to the other comparisons;


```{r, warning=FALSE, message=FALSE}
ggplot(data=plot_data, mapping = aes(x = Cell_Size_Uniformity, y = Bare_Nuclei)) +
  geom_point(aes(col=Class)) +
  geom_smooth(method = "lm", se = T) +
  labs(y="Nuclei With Cytoplasm", x="Cell Size Uniformity", 
       caption="Figure 5: Plot of the correlation between nuclei with cytoplasm and cell size uniformity.") +
  theme_minimal() +
  theme(plot.caption.position = "plot",
        plot.caption = element_text(hjust = 0.5))
```


```{r, warning=FALSE, message=FALSE}
ggplot(data=plot_data, mapping = aes(x = Cell_Shape_Uniformity, y = Bare_Nuclei)) +
  geom_point(aes(col=Class)) +
  geom_smooth(method = "lm", se = T) +
  labs(y="Nuclei With Cytoplasm", x="Cell Shape Uniformity", 
       caption="Figure 6: Plot of the correlation between nuclei with cytoplasm and cell shape uniformity.") +
  theme_minimal() +
  theme(plot.caption.position = "plot",
        plot.caption = element_text(hjust = 0.5))
```

The features that correlate less with the malignancy also seem to correlate slightly less amongst themselves, but let us make a heatmap of the entire dataset, to make sure we did not miss any other highly correlating attributes with our 80% cut-off.

```{r, warning=FALSE, message=FALSE}
cor_matrix <- round(cor(na.omit(data[,-1])),2)
melted_cor_matrix <- melt(cor_matrix)

ggplot(data = melted_cor_matrix, aes(x=Var1, y=Var2, fill=value, )) + 
  geom_tile() +
  labs(y="", x="", 
       caption="Figure 7: Heatmap showing the correlations between all attributes.") +
  scale_fill_gradient(low="black", high="red") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 40, hjust = 1)) +
  theme(plot.caption.position = "plot",
        plot.caption = element_text(hjust = 0.5))
```
We can clearly see cell uniformity and bare nuclei at the top, but it looks like all features correlate at least somewhat, with bland chromatin seemingly at the top of the sub-80% attributes with a correlation to class of roughly 76%. The heatmap also confirms the correlation between cell size and shape uniformity.

Having learnt this, let us now also make a graph of the spread for bland chromatin.

```{r, warning=FALSE, message=FALSE}
ggplot(data=plot_data, mapping = aes(x = Class, y = Bland_Chromatin)) +
  geom_jitter(width = 0.05, height = 0.3, col="cyan", alpha = 0.3) +
  labs(y="Bland Chromatin", 
       caption="Figure 8: Bland Chromatin spread per class.") +
  theme_minimal() +
  theme(plot.caption.position = "plot",
        plot.caption = element_text(hjust = 0.5))
```

This seems to somewhat mimic the spreads of the top 3 correlating features, albeit to a slightly lesser degree. Take note that a grade of 10 on the bland chromatin attribute means that the chromatin is maximally coarse and a 1 stands for very finely textured chromatin.


